Development and assessment of automated forest road projection methods using performance metrics relevant for wildlife

### 2024-06-11 analysis/paper/paper.Rmd
finish updating for new results and incorporate JH edits

We chose the Revelstoke timber supply area (TSA 27) in British Columbia, Canada as an assessment area because it is mountainous (Figure \@ref(fig:study-area)), containing approximately 38% non-forested land [@british_columbia_ministry_of_forests_revelstoke_2022] where road building is difficult or impossible. Thus, the area provides opportunities to observe the behaviour of road projection algorithms in steep complex terrain. Forestry is the dominant resource extraction activity and much of the road network was built to access forest harvest cutblocks, allowing projection of the road network from historic forest harvest data. The Revelstoke TSA includes portions of the Monashee and Selkirk mountain ranges, and is bisected by the Columbia River. At low to mid-elevations, forests of the interior cedar-hemlock biogeoclimatic zone [@british_columbia_ministry_of_forests_revelstoke_2022] include remnants of rare old growth inland temperate rainforest [@government_of_canada_parks_canada_agency_old_2017]. At higher elevations, forests are dominated by Engelmann spruce and subalpine fir [@british_columbia_ministry_of_forests_revelstoke_2022]. The area is within the ancestral territories of the Ktunaxa Nation, Secwepemc Nation, Syilx and Sinixt [@british_columbia_ministry_of_forests_revelstoke_2022]. Two of six southern mountain woodland caribou herds in the area have been extirpated, and the remaining herds are considered Endangered by the Committee on the Status of Endangered Wildlife in Canada and listed as threatened under the Canadian Species at Risk Act [@palm_long_2020; @british_columbia_ministry_of_forests_revelstoke_2022; @environment_canada_recovery_2014]. Mountain caribou avoid roads leading to a loss of available habitat [@polfus_identifying_2011; @apps_factors_2006], suffer increased mortality near roads [@apps_spatial_2013], and decline in abundance as the footprint of roads and cutblocks increases [@lochhead_linking_2022]. Roads also alter grizzly bear habitat use, home range selection, movements, population fragmentation, survival, reproductive rates, population density, trends, and conservation status [@proctor_effects_2020].

@@

According to all aspatial performance measures, the grade penalty method for setting edge weights yields better results than our simple cost methods (compare Figure \@ref(fig:aspat-perf-fig) and Figure B.1). We therefore recommend using the grade penalty method, and focus on further examining variation in performance among grade penalty method variants (Figure \@ref(fig:aspat-perf-fig)). The MST regular high density method yields the best projections of road density overall, while the MST random high density method had the best projections within cutblocks, and all MST variants yield comparably good projections outside cutblocks. All MST variants yield better projections of the forestry disturbance footprint than the Hardy or ILCP methods, and the Hardy and MST regular high density methods yield comparably accurate projections of road presence and distance to road.

All of the road projection methods underestimate, road presence and forestry disturbance footprint (Figure \@ref(fig:aspat-perf-fig)). In other words, these road network projection methods project fewer roads than we observe. Projections of distance to road, the forestry disturbance footprint, and the road disturbance footprint were more accurate than projections of road density and road presence (Figure \@ref(fig:aspat-perf-fig)), suggesting that these measures were more robust to our coarse assumptions about road projection. One reason for underestimation of road density is that all these methods generate straight road segments within a pixel. As raster resolution increases, the discrepancy within pixels between real roads and straight projected segments increases, which explains why projections of road density are less accurate when the raster resolution is coarse (Figure D.1).

@@

@@

@@

Although the MST method performed better than the alternatives we tested in terms of several important wildlife habitat metrics (Figure \@ref(fig:aspat-perf-fig)), it requires greater computation resources for large landscapes (Figure S7). All of these methods require construction of a large mathematical graph with a node for each pixel in the weight raster where road construction is possible (i.e. not NA), and as such, memory requirements will increase with the number of nodes in the graph (Figure S7). Compute time also increases with the number of landing targets (Figure S7). Lochhead et al's [-@lochhead_demo_2018; -@lochhead_linking_2022] method of graph construction is more computationally efficient than the adjacency matrix approach used in the R package `gdistance` [@van_etten_r_2017] (Figure S8), in part because it avoids the additional cost of constructing a transition matrix, and in part because it ignores variation in the size of raster cells that is accounted for in the `gdistance` _geoCorrection_ function. The Hardy QGIS plugin is 65 times faster than the MST and ILCP methods implemented in R (Figure S7).

@@

Among the method variants we considered, the minimum spanning tree (MST) regular high landing density method yielded the best projections of average road density and forestry disturbance footprint (Figure 3). Aspatial summaries of road presence, distance to road, and road disturbance footprint were comparable to Hardy's method [-@hardy_landis-ii_2023] (Figure 3). However, the MST algorithm was also more computationally demanding than the other methods we tested (Figure S7), so users will need to consider whether the benefits of improved performance outweigh the increased computational costs in their particular circumstance.

Aspatial metrics of overall projection accuracy can be difficult to interpret. To more deeply understand variation in the behaviour of road network projection algorithms we found it helpful to distinguish between performance within cutblocks and outside of them, and to examine spatial patterns. This examination highlighted some errors that could be fixed by better accounting for the cost of water crossings in our weight raster, and other more fundamental limitations and differences among methods. Even in cases where the Hardy and grade penalty methods project similar mean road densities, the projected pattern of roads within cutblocks is quite different, in that the grade penalty roads tend to follow elevation contours, while Hardy et al's method produces curves (Figure 4). We suspect the latter behaviour is an artifact of searching for nearest destinations \>200 m away on a 100 m grid.

@@

These road projection methods require construction of large graphs, with a node for each pixel in which road construction is possible. While we were able to run the models in our example landscape up to 100 m resolution in a reasonable amount of time (< 2.4 hours) using modest amounts of memory (<3 GB RAM), projections at 30 m and 10 m resolutions caused R to crash due to a lack of available memory (Supplement E). Although this challenge can be fairly easily overcome by running the projections on a machine with more RAM, the memory and computational requirements may present a challenge on larger landscapes or when incorporated into more complex simulation models. For example, for a project studying the effects of different forest management and climate change scenarios on wildlife in the Churchill caribou range in Ontario, we projected road development across the landscape, which at 50 m resolution has a graph with greater than 21 million nodes. Projecting roads using the ILCP method on this landscape took 2.7 hours to run and required 11 GB of RAM for one realization. Overall we hope to complete 600 different realizations which will require significant time and resources.

Options for reducing computational requirements include restricting the area where road construction is possible (by setting some portions of the weight raster to NA), reducing landing density within cutblocks (by using the centroid method), increasing raster resolution, or using Hardy's more efficient implementations. The first three options are available to users of the `roads` R package. However, those contemplating reducing computational cost by decreasing raster resolution should be cautious of the results given that all of these methods become less accurate as raster cell size increases (Figures S5 and S6). The distance to road metric is less sensitive to changes in cell resolution, and distance metrics may also be more ecologically meaningful [@rowland_elk_2000]. Those considering reducing computational cost by reducing landing density within cutblocks should be clear that their projected road network will be incomplete and metrics such as total road length or road density will be underestimated, while the combined footprint of roads and cutblocks should be reasonably accurate. More labour intensive possibilities for reducing computational costs include implementing slow components of the R package (Figure S8) in a faster compiled language, and implementing a multi-scale approach in which coarse resolution results are used to limit the area considered in finer resolution projections.

There are many metrics of connectivity and fragmentation [@francis_routledge_2021] we did not consider in this paper. In general, performance will vary among metrics, and it is difficult to determine what is good enough without reference to particular circumstances (i.e. objectives, species, landscapes). For example, where the outcome of interest is space use or connectivity projected from more [e.g. @whittington_towns_2022; @labadie_global_2023] or less [e.g. @diniz_landscape_2020; @hughes_comparison_2023] species-specific movement models it is recommended to test their sensitivity to the difference between observed and retrospectively projected roads. When constructing new wildlife response models intended for projection, it may also be helpful to select metrics such as distance to road or buffered disturbance footprint that are more easily projectable [@bodner_bridging_2021]. In lieu of testing additional performance metrics, we have focused our efforts on the development of an open source R package [@endicott_roads_2023] and workflows [@endicott_roadpaper_2024] that enable users to devise performance tests appropriate to their particular circumstances and goals. We recognize that acquiring data to retrospectively project roads can be challenging, so we also provide output maps [@endicott_assessment_2024] and an example script (analysis/scripts/4_calc_any_metric.R) to enable calculation of alternative performance metrics of interest for our Revelstoke example, along with a data preparation script (analysis/scripts/1_prepareData.R) that could be modified to extract required input data for other areas in British Columbia [@endicott_roadpaper_2024].

@@

\newpage
# Supplementary Material

@@

\newpage
# Figures

@@

\newpage
Figure 2: Proportional difference of projected mean from observed mean road metrics using different projection method variants within cutblocks, outside cutblocks, and overall (1 ha weight raster). Values greater than zero indicate overprojection and negative values indicate underprojection. Cutblocks not accessible via the observed road network were excluded from analysis. Performance varies among metrics (columns), projection method variants (colours), and within and outside cutblocks. Projection method variants are defined in Table 2.

```{r aspat-perf-fig, fig.height=4, fig.width=fig_widths["two"], out.width=paste0(fig_widths["two"], "in"), out.height="4in", fig.cap="."}
# organize data for ggplot
doAspatPlot(meanTableRealCuts)
```

@@

Figure 4: Comparison of the minimum spanning tree projection with regular high density sampling to the observed road network. The main map shows observed distance to the nearest road minus the projected distance. Insets highlight areas of interest with satellite imagery (Google 2022) for context: (a) Projected and observed road networks follow valleys but projected roads cross valley bottoms more frequently; (b) some open areas are misidentified as cutblocks; (c) projected roads follow contours but do not avoid turns and dead ends like observed switchbacks.
```{r proj-map-fig, fig.height=9, fig.width=fig_widths["two"], fig.cap="."}

@@

Table 1: Road construction costs. In the “grade penalty” method, we assign grades >20% a more prohibitive cost of 65000, which is also assigned to major water bodies. These barriers are not set to NA because roads must cross barriers to reach a few isolated cutblocks in our example landscape.
```{r cost-tbl}
tbl <- "Cost Type|Value ($ per km)|Source
Slope (%)|16178 + 504*slope|Interior appraisal manual
Major lakes|65000|Major water bodies of BC
Existing roads|0|Digital road atlas and forest tenure roads for BC
"
df <- read.delim(textConnection(tbl),header=FALSE,sep="|",strip.white=TRUE,stringsAsFactors=FALSE)
names(df) <- unname(as.list(df[1,]))
df <- df[-1,] # remove first row
row.names(df)<-NULL
knitr::kable(df, caption = ".")
```

@@

Table 2: Road projection method variants. We examined sensitivity of minimum spanning tree (MST) and iterative least cost path (ILCP) algorithms to the method of setting edge weights (grade penalty or simple cost), the placement of landings within cutblocks (regular or random), the density of landings within cutblocks, and map resolution.
```{r method-tbl}
tbl2 <- "Projection method|Sample type|Sample density (pts/km^2^)|Resolutions
MST grade penalty & simple cost|Regular|1|1 & 100 ha
MST grade penalty & simple cost|Regular|10|1 & 100 ha
MST grade penalty & simple cost|Random|1|1 & 100 ha
MST grade penalty & simple cost|Random|10|1 & 100 ha
MST grade penalty & simple cost|Centroid|N/A|1 & 100 ha
ILCP grade penalty & simple cost|Regular|10|1 ha 
Hardy QGIS|N/A|~9|1 ha
Cutblocks only|N/A|N/A|1 ha

@@

\newpage
### Colophon

>>>>

We chose the Revelstoke timber supply area (TSA 27) in British Columbia, Canada as an assessment area because it is mountainous (Figure \@ref(fig:study-area)), containing approximately 38% non-forested land [@british_columbia_ministry_of_forests_revelstoke_2022] where road building is challenging. Thus, the area provides opportunities to observe the behaviour of road projection algorithms in steep complex terrain. Forestry is the dominant resource extraction activity and much of the road network was built to access forest harvest cutblocks, allowing projection of the road network from historic forest harvest data. The Revelstoke TSA includes portions of the Monashee and Selkirk mountain ranges, and is bisected by the Columbia River. At low to mid-elevations, forests of the interior cedar-hemlock biogeoclimatic zone [@british_columbia_ministry_of_forests_revelstoke_2022] include remnants of rare old growth inland temperate rainforest [@government_of_canada_parks_canada_agency_old_2017]. At higher elevations, forests are dominated by Engelmann spruce and subalpine fir [@british_columbia_ministry_of_forests_revelstoke_2022]. The area is within the ancestral territories of the Ktunaxa Nation, Secwepemc Nation, Syilx and Sinixt [@british_columbia_ministry_of_forests_revelstoke_2022]. Two of six southern mountain woodland caribou herds in the area have been extirpated, and the remaining herds are considered Endangered by the Committee on the Status of Endangered Wildlife in Canada and listed as threatened under the Canadian Species at Risk Act [@palm_long_2020; @british_columbia_ministry_of_forests_revelstoke_2022; @environment_canada_recovery_2014]. Mountain caribou avoid roads leading to a loss of available habitat [@polfus_identifying_2011; @apps_factors_2006], suffer increased mortality near roads [@apps_spatial_2013], and decline in abundance as the footprint of roads and cutblocks increases [@lochhead_linking_2022]. Roads also alter grizzly bear habitat use, home range selection, movements, population fragmentation, survival, reproductive rates, population density, trends, and conservation status [@proctor_effects_2020].

@@

@@

aspat_perf <- doAspatPlot(meanTableRealCuts)

@@

mst_reg_dens <- aspat_perf$prop_dif %>% 
  filter(metric == "Road\ndensity", sampleDens == "MST regular high density", 
         areaMean == "Overall")

mst_ran_dens <- aspat_perf$prop_dif %>% 
  filter(metric == "Road\ndensity", sampleDens == "MST random high density", 
         areaMean == "Cutblocks")

mst_out_dens <- aspat_perf$prop_dif %>% 
  filter(metric == "Road\ndensity", method == "mst", areaMean == "Outside\ncutblocks")

all_over_fordens <- aspat_perf$prop_dif %>% 
  filter(metric == "Forestry\ndisturbance\nfootprint", areaMean == "Overall")
```

According to all aspatial performance measures, the grade penalty method for setting edge weights yields better results than our simple cost methods (compare Figure \@ref(fig:aspat-perf-fig) and Figure S1). We therefore recommend using the grade penalty method, and focus on further examining variation in performance among grade penalty method variants (Figure \@ref(fig:aspat-perf-fig)). The MST regular high density method yields the best projections of road density overall (projected density: `r mst_reg_dens$response_proj %>% round(2)` m roads/ha, observed density: `r mst_reg_dens$response_obs %>% round(2)` m roads/ha), while the MST random high density method had the best projections within cutblocks (projected density: `r mst_ran_dens$response_proj %>% round(2)` m roads/ha, observed density: `r mst_ran_dens$response_obs %>% round(2)` m roads/ha), and all MST variants yield comparably good projections outside cutblocks (projected density: `r mst_out_dens$response_proj %>% range() %>% round(2) %>% paste0(collapse = " - ")` m roads/ha, observed density: `r mst_out_dens$response_obs %>% unique() %>% round(2)` m roads/ha). All MST variants yield better projections of the forestry disturbance footprint (projected: `r all_over_fordens %>% filter(method == "mst") %>% pull(response_proj) %>% range() %>% round(2) %>% unique()`, observed: `r all_over_fordens %>% filter(method == "mst") %>% pull(response_obs) %>% unique() %>% round(2)`) than the Hardy or ILCP methods (`r all_over_fordens %>% filter(sampleDens %in% c("Hardy QGIS", "ILCP regular high density")) %>% pull(response_proj) %>% round(2) %>% unique()`), and the Hardy and MST regular high density methods yield comparably accurate projections of road presence and distance to road.

All of the road projection methods underestimate, road presence and forestry disturbance footprint (Figure \@ref(fig:aspat-perf-fig)). In other words, these road network projection methods project fewer roads than we observe. Projections of the forestry disturbance footprint, the road disturbance footprint, and, to a lesser extent, distance to road were more accurate than projections of road density and road presence (Figure \@ref(fig:aspat-perf-fig)), suggesting that these measures were more robust to our coarse assumptions about road projection. One reason for underestimation of road density is that all these methods generate straight road segments within a pixel. As raster resolution increases, the discrepancy within pixels between real roads and straight projected segments increases, which explains why projections of road density are less accurate when the raster resolution is coarse (Figure S6).

@@

Although the MST method performed better than the alternatives we tested in terms of several important wildlife habitat metrics (Figure \@ref(fig:aspat-perf-fig)), it requires greater computation resources for large landscapes (Figure S7). All of these methods require construction of a large mathematical graph with a node for each pixel in the weight raster where road construction is possible (i.e. not NA), and as such, memory requirements will increase with the number of nodes in the graph (Figure S7). Compute time also increases with the number of landing targets (Figure S7). Lochhead et al's [-@lochhead_demo_2018; -@lochhead_linking_2022] method of graph construction is more computationally efficient than the adjacency matrix approach used in the R package `gdistance` [@van_etten_r_2017] (Figure S8), in part because it avoids the additional cost of constructing a transition matrix, and in part because it ignores variation in the size of raster cells that is accounted for in the `gdistance` *geoCorrection* function. The Hardy QGIS plugin is 65 times faster than the MST and ILCP methods implemented in R (Figure S7).

@@

Among the method variants we considered, the minimum spanning tree (MST) regular high landing density method yielded the best projections of average road density and forestry disturbance footprint (Figure \@ref(fig:aspat-perf-fig)). Aspatial summaries of road presence, distance to road, and road disturbance footprint were comparable to Hardy's method [-@hardy_landis-ii_2023] (Figure \@ref(fig:aspat-perf-fig)). However, the MST algorithm was also more computationally demanding than the other methods we tested (Figure S7), so users will need to consider whether the benefits of improved performance outweigh the increased computational costs in their particular circumstance.

Aspatial metrics of overall projection accuracy can be difficult to interpret. To more deeply understand variation in the behaviour of road network projection algorithms we found it helpful to distinguish between performance within cutblocks and outside of them, and to examine spatial patterns. This examination highlighted some errors that could be fixed by better accounting for the cost of water crossings in our weight raster, and other more fundamental limitations and differences among methods. Even in cases where the Hardy and grade penalty methods project similar mean road densities, the projected pattern of roads within cutblocks is quite different, in that the grade penalty roads tend to follow elevation contours, while Hardy et al's method produces curves (Figure \@ref(fig:proj-meth-fig)). We suspect the latter behaviour is an artifact of searching for nearest destinations \>200 m away on a 100 m grid.

@@

These road projection methods require construction of large graphs, with a node for each pixel in which road construction is possible. While we were able to run the models in our example landscape up to 100 m resolution in a reasonable amount of time (\< 2.4 hours) using modest amounts of memory (\<3 GB RAM), projections at 30 m and 10 m resolutions caused R to crash due to a lack of available memory (Supplement E). Although this challenge can be fairly easily overcome by running the projections on a machine with more RAM, the memory and computational requirements may present a challenge on larger landscapes or when incorporated into more complex simulation models. For example, for a project studying the effects of different forest management and climate change scenarios on wildlife in the Churchill caribou range in Ontario, we projected road development across the landscape, which at 50 m resolution has a graph with greater than 21 million nodes. Projecting roads using the ILCP method on this landscape took 2.7 hours to run and required 11 GB of RAM for one realization. Overall we hope to complete 600 different realizations which will require significant time and resources.

Options for reducing computational requirements include restricting the area where road construction is possible (by setting some portions of the weight raster to NA), reducing landing density within cutblocks (by using the centroid method), increasing raster resolution, or using Hardy's more efficient implementations. The first three options are available to users of the `roads` R package. However, those contemplating reducing computational cost by decreasing raster resolution should be cautious of the results given that all of these methods become less accurate as raster cell size increases (Figures S5 and S6). The distance to road metric is less sensitive to changes in cell resolution, and distance metrics may also be more ecologically meaningful [@rowland_elk_2000]. Those considering reducing computational cost by reducing landing density within cutblocks should be clear that their projected road network will be incomplete and metrics such as total road length and road density will be underestimated. More involved possibilities for reducing computational costs include adopting faster alternative routing algorithms [@larmet_cppRouting_2022], implementing slow components of the R package (Figure S8) in a faster compiled language, parallel processing, and implementing a multi-scale approach in which coarse resolution results are used to limit the area considered in finer resolution projections. Several of these options have been implemented in the cppRouting R package [@larmet_cppRouting_2022], and future versions of the roads R package could borrow and build on cppRouting functionality.

There are many metrics of connectivity and fragmentation [@francis_routledge_2021] we did not consider in this paper. In general, performance will vary among metrics, and it is difficult to determine what is good enough without reference to particular circumstances (i.e. objectives, species, landscapes). For example, if the outcome of interest is space use or connectivity projected from more [e.g. @whittington_towns_2022; @labadie_global_2023] or less [e.g. @diniz_landscape_2020; @hughes_comparison_2023] species-specific movement models it may be helpful to test sensitivity to the difference between observed and retrospectively projected roads, in order to confirm that projections are adequate for the intended purpose. When constructing new wildlife response models intended for projection, it may also be helpful to select metrics such as distance to road or buffered disturbance footprint that are more easily projectable [@bodner_bridging_2021]. In lieu of testing additional performance metrics, we have focused our efforts on the development of an open source R package [@endicott_roads_2023] and workflows [@endicott_roadpaper_2024] that enable users to devise performance tests appropriate to their particular circumstances and goals. We recognize that acquiring data to retrospectively project roads can be challenging, so we also provide output maps [@endicott_assessment_2024] and an example script (analysis/scripts/4_calc_any_metric.R) to enable calculation of alternative performance metrics of interest for our Revelstoke example, along with a data preparation script (analysis/scripts/1_prepareData.R) that could be modified to extract required input data for other areas in British Columbia [@endicott_roadpaper_2024].

@@

# Supplementary Material

@@

# Figures

@@

Figure 2: Proportional difference of projected mean from observed mean road metrics using different projection method variants within cutblocks, outside cutblocks, and overall (1 ha weight raster). Values greater than zero indicate overprojection and negative values indicate underprojection. Cutblocks not accessible via the observed road network were excluded from analysis. Performance varies among metrics (columns), projection method variants (colours), and within and outside cutblocks. Projection method variants are defined in Table 2. Note that extreme values for the cutblocks only method within cutblocks are cut off to ensure patterns in the other methods to remain visible. Bars that are cutoff are labeled with the total value.

```{r aspat-perf-fig, fig.height=4, fig.width=fig_widths["two"], out.width=paste0(fig_widths["two"], "in"), out.height="4in", fig.cap="."}
# organize data for ggplot
# need to make plots more comparable
# Other option is setting limits but want to avoid excess white space
ann_plus <- aspat_perf$prop_dif %>% 
  filter(abs(prop_dif) > 0.5) %>% 
  mutate(lab = round(prop_dif, 2), 
         prop_dif = 0.37*sign(prop_dif))

aspat_perf$plt+coord_flip(ylim = c(-0.35, 0.35))+
  geom_text(aes(label = lab), data = ann_plus, hjust = "inward", vjust = 0.5,
            size = 8, size.unit = "pt")
```

@@

Figure 4: Comparison of the minimum spanning tree projection with regular high density sampling to the observed road network. The main map shows observed distance to the nearest road minus the projected distance. Insets highlight areas of interest with satellite imagery (Google 2022) for context: (a) Projected and observed road networks follow valleys but projected roads cross valley bottoms more frequently; (b) some open areas are misidentified as cutblocks; (c) projected roads follow contours but do not avoid turns and dead ends like observed switchbacks.

```{r proj-map-fig, fig.height=9, fig.width=fig_widths["two"], fig.cap="."}

@@

Table 1: Road construction costs. In the “grade penalty” method, we assign grades \>20% a more prohibitive cost of 65000, which is also assigned to major water bodies. These barriers are not set to NA because roads must cross barriers to reach a few isolated cutblocks in our example landscape.

```{r cost-tbl}
tbl <- "Cost Type|Value ($ per km)|Source
Slope (%)|16178 + 504*slope|Interior appraisal manual
Major lakes|65000|Major water bodies of BC
Existing roads|0|Digital road atlas and forest tenure roads for BC
"
df <- read.delim(textConnection(tbl),header=FALSE,sep="|",strip.white=TRUE,stringsAsFactors=FALSE)
names(df) <- unname(as.list(df[1,]))
df <- df[-1,] # remove first row
row.names(df)<-NULL
knitr::kable(df, caption = ".")
```

@@

Table 2: Road projection method variants. We examined sensitivity of minimum spanning tree (MST) and iterative least cost path (ILCP) algorithms to the method of setting edge weights (grade penalty or simple cost), the placement of landings within cutblocks (regular or random), the density of landings within cutblocks, and map resolution.

```{r method-tbl}
tbl2 <- "Projection method|Sample type|Sample density (pts/km^2^)|Resolutions
MST grade penalty & simple cost|Regular|1|1 & 100 ha
MST grade penalty & simple cost|Regular|10|1 & 100 ha
MST grade penalty & simple cost|Random|1|1 & 100 ha
MST grade penalty & simple cost|Random|10|1 & 100 ha
MST grade penalty & simple cost|Centroid|N/A|1 & 100 ha
ILCP grade penalty & simple cost|Regular|10|1 ha 
Hardy QGIS|N/A|~9|1 ha
Cutblocks only|N/A|N/A|1 ha

@@

### Colophon

### 2024-06-11 analysis/paper/supplement/supplement.Rmd
finish updating for new results and incorporate JH edits

```{r aspat-perf-fig-simp, fig.height=4, fig.width=fig_widths["two"], fig.cap="Proportional difference of projected mean from observed mean road metrics using different simple cost projection method variants within cutblocks, outside cutblocks, and overall (1 ha cost raster). Values greater than zero indicate overprojection and negative values indicate underprojection. Cutblocks not accessible via the observed road network were excluded from analysis. Performance varies among metrics (columns), projection method variants (colours), and within and outside cutblocks."}
# organize data for ggplot
doAspatPlot(meanTableRealCuts)
```

@@

```{r spat-perf-fig-simp, fig.height=4.5, fig.width=fig_widths["two"], fig.cap="Spatial variation of in performance among simple cost projection method variants (1 ha cost raster). F-measure is the harmonic mean of precision and sensitivity, precision is the proportion of predicted pixels that were observed and sensitivity is the proportion of observed pixels that were correctly predicted. Values closer to one indicate better performance. Cutblocks not accessible via the observed road network were excluded from analysis."}
perf_tbl <- doSpatPerf(matchDataRealCuts)
perf_tbl_rng <- perf_tbl %>% group_by(metric) %>% 
  summarise(rng_F = range(F_measure, na.rm = TRUE) %>% round(3) %>% paste0(collapse = " - "), .groups = "drop") %>% pull(rng_F, name = metric)

doSpatPlot(perf_tbl)

@@

```{r aspat-perf-fig-all, fig.height=4, fig.width=fig_widths["two"], fig.cap="Proportional difference of projected mean from observed mean road metrics using different projection method variants with a 1 ha resolution weight raster in analysis that includes false cutblocks with no observed access roads. Metrics were summarised within cutblocks, outside cutblocks, and overall. Values greater than 0 indicate overprojection of the metric and negative values indicate an underprojection. "}
meanTable <- read.csv(here(data_path_drvd, "dem", "mean_table.csv"))
doAspatPlot(meanTable)
```

@@

```{r figA1, fig.height=6, fig.width=fig_widths["two"], fig.cap="Proportional difference of projected mean from observed mean road metrics using different projection methods for the fine resolutiom (1 ha) and coarse resolution (100 ha) weight rasters. Metrics were summarised within cutblocks, outside cutblocks, and overall. Values greater than 0 indicate the projection overestimated the metric and negative values indicate an underestimation."}
# organize data for ggplot
meanTable_long <- meanTable %>% dplyr::select(-runTime, -order) %>% 
  pivot_longer(-c(sampleType, sampleDens, areaMean, resolution, method),
               names_to = "metric", values_to = "response")%>%
  mutate(sampleType = ifelse(sampleType == sampleDens, NA_character_, 
                             sampleType),
         sampleDens = paste(method, sampleType, sampleDens) %>% 
           factor(levels = c("mst NA centroid",
                             "NA NA observed",
                             "mst random low sample density",
                             "mst regular low sample density",
                             "mst random high sample density",
                             "mst regular high sample density",
                             "ilcp regular high sample density",
                             "NA klementQGIS", 
                             "NA NA NA"),
                  labels = c("MST centroid", 
                             "Observed",
                             "MST random low density",
                             "MST regular low density",
                             "MST random high density",
                             "MST regular high density",
                             "ILCP regular high density",
                             "Hardy QGIS",
                             "Cutblocks only")),
         metric = factor(metric,
                         levels = c("roadDensityMean", "roadPresenceMean",
                                    "distanceToRoadMean",
                                    "forestryDisturbanceMean",
                                    "roadDisturbanceMean"),
                         labels = c("Road\ndensity", "Road\npresence",
                                    "Distance\nto road",
                                    "Forestry\ndisturbance\nfootprint",
                                    "Road\ndisturbance\nfootprint")))

>>>>

```{r aspat-perf-fig-simp, fig.height=4, fig.width=fig_widths["two"], fig.cap="Proportional difference of projected mean from observed mean road metrics using different simple cost projection method variants within cutblocks, outside cutblocks, and overall (1 ha cost raster). Values greater than zero indicate overprojection and negative values indicate underprojection. Cutblocks not accessible via the observed road network were excluded from analysis. Performance varies among metrics (columns), projection method variants (colours), and within and outside cutblocks."}
# organize data for ggplot
aspat_simp <- doAspatPlot(meanTableRealCuts)

ann_plus <- aspat_simp$prop_dif %>% 
  filter(abs(prop_dif) > 0.5) %>% 
  mutate(lab = round(prop_dif, 2), 
         prop_dif = 0.37*sign(prop_dif))

aspat_simp$plt+
  coord_flip(ylim = c(-0.35, 0.35))+
  geom_text(aes(label = lab), data = ann_plus, hjust = "inward", vjust = 0.5,
            size = 10, size.unit = "pt")

@@

```{r spat-perf-fig-simp, fig.height=4.5, fig.width=fig_widths["two"], fig.cap="Spatial variation of in performance among simple cost projection method variants (1 ha cost raster). F-measure is the harmonic mean of precision and sensitivity, precision is the proportion of predicted pixels that were observed and sensitivity is the proportion of observed pixels that were correctly predicted. Values closer to one indicate better performance. Cutblocks not accessible via the observed road network were excluded from analysis."}
perf_tbl_simp <- doSpatPerf(matchDataRealCuts)
perf_tbl_rng_simp <- perf_tbl_simp %>% group_by(metric) %>% 
  summarise(rng_F = range(F_measure, na.rm = TRUE) %>% round(3) %>% paste0(collapse = " - "), .groups = "drop") %>% pull(rng_F, name = metric)

doSpatPlot(perf_tbl_simp)

@@

```{r aspat-perf-fig-all, fig.height=4, fig.width=fig_widths["two"], fig.cap="Proportional difference of projected mean from observed mean road metrics using different projection method variants with a 1 ha resolution weight raster in analysis that includes false cutblocks with no observed access roads. Metrics were summarised within cutblocks, outside cutblocks, and overall. Values greater than 0 indicate overprojection of the metric and negative values indicate an underprojection. Note that extreme values for the cutblocks only method within cutblocks are cut off to ensure patterns in the other methods to remain visible. Bars that are cutoff are labeled with the total value."}
meanTable <- read.csv(here(data_path_drvd, "dem", "mean_table.csv"))
aspat_all <- doAspatPlot(meanTable)

ann_plus <- aspat_all$prop_dif %>% 
  filter(abs(prop_dif) > 0.5) %>% 
  mutate(lab = round(prop_dif, 2), 
         prop_dif = 0.37*sign(prop_dif))

aspat_all$plt+
  coord_flip(ylim = c(-0.35, 0.35))+
  geom_text(aes(label = lab), data = ann_plus, hjust = "inward", vjust = 0.5,
            size = 10, size.unit = "pt")
```

@@

```{r figA1, fig.height=6, fig.width=fig_widths["two"], fig.cap="Proportional difference of projected mean from observed mean road metrics using different projection methods for the fine resolutiom (1 ha) and coarse resolution (100 ha) weight rasters. Metrics were summarised within cutblocks, outside cutblocks, and overall. Values greater than 0 indicate the projection overestimated the metric and negative values indicate an underestimation. Note that each panel has a different scale so bars cannot be compared across metrics."}
# organize data for ggplot
meanTable_long <- meanTable %>% dplyr::select(-runTime, -order) %>% 
  pivot_longer(-c(sampleType, sampleDens, areaMean, resolution, method),
               names_to = "metric", values_to = "response")%>%
  mutate(sampleType = ifelse(sampleType == sampleDens, NA_character_, 
                             sampleType),
         sampleDens = paste(method, sampleType, sampleDens) %>% 
           factor(levels = c("mst NA centroid",
                             "NA NA observed",
                             "mst random low sample density",
                             "mst regular low sample density",
                             "mst random high sample density",
                             "mst regular high sample density",
                             "ilcp regular high sample density",
                             "NA klementQGIS", 
                             "NA NA NA"),
                  labels = c("MST centroid", 
                             "Observed",
                             "MST random low density",
                             "MST regular low density",
                             "MST random high density",
                             "MST regular high density",
                             "ILCP regular high density",
                             "Hardy QGIS",
                             "Cutblocks only")),
         metric = factor(metric,
                         levels = c("roadDensityMean", "roadPresenceMean",
                                    "distanceToRoadMean",
                                    "forestryDisturbanceMean",
                                    "roadDisturbanceMean"),
                         labels = c("Road\ndensity", "Road\npresence",
                                    "Distance\nto road",
                                    "Forestry\ndisturbance\nfootprint",
                                    "Road\ndisturbance\nfootprint")))
